---
title: "Homework 3"
author: "Wayne Monical wem2121"
date: "2024-10-10"
output: github_document
---

## Problem 1

```{r}
library(tidyverse)
library(p8105.datasets)
data("ny_noaa")
```


The `ny_noaa` data set has `r ncol(ny_noaa)` columns and `r nrow(ny_noaa)` rows. Each observation comes with an ID given by the `id` variable. Each observation has a date, along with several precipitation and temperature measurements. 



We begin with data cleaning by creating separate numeric variables for year, month, and day. 

```{r}
ny_noaa=
  ny_noaa |> 
  mutate(
    year = ny_noaa |> pull(date) |> year(),
    month= ny_noaa |> pull(date) |> month(),
    day= ny_noaa |> pull(date) |> day()
  )
```

Using these new variables, we find that we have observations from `r min(pull(ny_noaa, year))` to `r max(pull(ny_noaa, year))`.

Looking at the data's description at https://p8105.com/dataset_noaa.html, we can find that the precipitation is given in tenths of millimeters, snowfall and snow depth are given in millimeters, and maximum and minimum temperature are given in tenths of degrees Celsius. Therefore we can adjust the columns to make each column in whole millimeters or whole degrees Celsius. 
```{r}
ny_noaa = 
  ny_noaa |> 
  mutate(
    prcp = ny_noaa$prcp / 10,
    tmin = as.numeric(ny_noaa$tmin) / 10,
    tmax = as.numeric(ny_noaa$tmax) / 10)
```

There is also a single observation with negative snowfall, with ID USC00307400 on June 15th, 2005. 
```{r}
ny_noaa |> filter(snow < 0)
```

Given that New York typically does not receive snowfall in June, this value is likely meant to be zero. However, the temperature values are both within two standard deviations of the average temperature for June, so I do not believe that this observation need be thrown out. 

```{r}
ny_noaa |> 
  filter(month == 6) |> 
  drop_na(tmin, tmax) |> 
  summarise(
    mean_june_tmin = mean(tmin),
    sd_june_tmin = sd(tmin),
    mean_june_tmax = mean(tmax),
    sd_june_tmax = sd(tmax))
```
```{r}
ny_noaa =
  ny_noaa |> 
  mutate(snow = ifelse(id == 'USC00307400', 0, snow))
```


However, even after correcting the units to millimeters, the daily precipitation, snowfall, and snow depth have overly large maximum values. The maximum snowfall for New York City was recorded as 693 millimeters on January 23, 2016 according to [this website](https://www.currentresults.com/Yearly-Weather/USA/NY/New-York-City/extreme-annual-new-york-city-snowfall.php). I do not see a clear cause of this issue. Errors in data collection, such as the negative snowfall variable, could be at fault. I would like to ask the data collector about these values. If I could not do that, I would simply discard any values over NYC's recorded maximum. 
```{r}
ny_noaa |> 
  select(prcp, snow, snwd) |> 
  summary()
```


The most common values for snowfall are zero, 25, 13, and 51. I suspect that this is because they are converting from inches and rounding to the nearest millimeter. This is because there are 25.4 millimeters in an inch. So half of one inch corresponds to 13 millimeters, one inch corresponds to 25 millimeters, and two inches corresponds to 51 millimeters, which we see are the most common values. 

```{r}
ny_noaa |> 
  count(snow) |> 
  arrange(desc(n)) |> 
  head(5)
```
Below I have made two graphs comparing the daily maximum temperature to the minimum temperature. The first graph is a two dimensional histogram. We can see that there may be a positive relationship between minimum and maximum temperature. 

```{r}
ny_noaa |> 
  drop_na(tmin, tmax) |> 
  ggplot(aes(x = tmin, y = tmax)) +
  geom_bin_2d()+ 
  labs(title = 'NYC Minimum and Maximum Daily Temperatures')+
  xlab('Minimum Daily Temperature') +
  ylab('Maximum Daily Temperature')
```

The next graph is a histogram faceted based on temperature statistic: minimum or maximum. It allows us to compare the two populations, but we have lost the information on which readings are on the same day. It is worth noting that these distributions are not close to normal. 
```{r}
ny_noaa |> 
  drop_na(tmin, tmax) |> 
  pivot_longer(
    cols = c('tmin', 'tmax'),
    names_to = 'temp_stat',
    values_to = 'temp_value') |> 
  ggplot(aes(x = temp_value)) +
  facet_grid(temp_stat~.)+
  geom_histogram()+ 
  labs(title = 'NYC Minimum and Maximum Temperatures')+
  xlab('Temperature in Degrees Celsius') +
  ylab('Count of Observations')
```


Next we create a plot showing the distribution of snowfall values greater than 0 millimeters and less than 100 millimeters in yeas of 2001 to 2004. As we previously observed, the measurements are biased towards increments of one-quarter inches. Therefore, I have specified the histogram's bins to match. 

```{r}
ny_noaa |> 
  filter(
    snow > 0,
    snow < 100,
    year %in% c(2001, 2002, 2003, 2004)
    ) |> 
  ggplot(aes(x = snow)) + 
  geom_histogram(binwidth = 25) + 
  facet_grid(rows = year~.)+ 
  labs(title = 'NYC Snowfall by Year')+
  xlab('Snowfall in Millimeters') +
  ylab('Count of Observations')
```


## Problem 2

### Loading and Cleaning Data

Here we load the NHANES acceleration data. The acceleration data comes in a wide format. We pivot the data to be longer in order to tidy it. We also transform the minute data to be numeric. We import the covariates data set and find that it is already tidy. 
```{r}
accel = 
  read_csv('data/nhanes_accel.csv') |> 
  janitor::clean_names() |> 
  pivot_longer(
    cols = min1:min1440,
    names_to = 'minute',
    names_prefix = 'min',
    values_to = 'acceleration'
  ) |> 
  mutate(minute = as.numeric(minute))

covar = 
  read_csv('data/nhanes_covar.csv', skip = 4) |> 
  janitor::clean_names()
```


We check for missing data, and find the ID for each individual, `seqn`, agrees. 
```{r}
anti_join(covar, accel)
```

Finally, we join together the data, and filter for participants of at least 21 years of age, full demographic data, and encode the qualitative variables as factors of character strings.

```{r}
nhanes = 
  inner_join(covar, accel) |> 
  filter(age >= 21) |> 
  drop_na(sex, age, bmi, education) |> 
  mutate(
    sex = 
      factor(
        case_when(
          sex == 1 ~ 'Male', 
          sex == 2 ~ 'Female',
          !(sex %in% c(1,2)) ~ NA 
        ),
        levels = c('Male', 'Female')
      ),
    education = 
      factor(
       case_when(
         education == 1 ~ "Less than High School",
         education == 2 ~ "High School Equivalent",
         education == 3 ~ "More than High School",
         !(education %in% c(1,2, 3)) ~ NA 
       ),
       levels = c(
         'Less than High School', 
         'High School Equivalent',
         "More than High School"
         )
        )
  )
```


### Making Tables

Next, we make a reader-friendly table comparing sex to educational attainment. We accomplish this by selecting for the relevant rows, dropping duplicate rows, grouping our data by sex and education, counting the observations, then formatting the table to be readable. From the table, we can see that there are approximately the same number of men and women in the study that have less than a high school education and more than a high school education. There are more men than women with a high school education. 

```{r}
nhanes |> 
  select(seqn, sex, education) |> 
  distinct() |> 
  group_by(sex, education) |> 
  summarize(count = n()) |>
  pivot_wider(
    names_from = sex, 
    values_from = count
    ) |> 
  knitr::kable()
```

Aggregating our data across all minutes in the day, we can plot the total activity separated by age, sex, and education. While we cannot draw any definitive conclusions without statistical tests, it appears that activity has a general downward relationship with age across sex and education. Among the high school educated and further, female study participants had an overall higher level of activity than their male counterparts. 
```{r}
nhanes |> 
  group_by(
    seqn, sex, age, education
  ) |> 
  summarize(total_activity = sum(acceleration)) |> 
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  facet_grid(.~education) +
  geom_point() +
  geom_smooth()+ 
  labs(title = 'Total Daily Acitivity by Age, Education, and Sex')
  
```

Aggregating instead by hour of the day, we see very similar activity levels across sex and education. Activity increases from the fifth hour to the tenth, where it gradually decreases until the twentieth hour, where it drops off steeply. This pattern is affirmed by humans' typically sleep cycle. Note that the graph below does not show all data points, as it has been rescaled for clarity. 

```{r}
nhanes |> 
  mutate(
    hour = floor(minute / 60)
  ) |> 
  group_by(
    seqn, sex, age, education, hour
  ) |> 
  summarize(hourly_activity = sum(acceleration)) |> 
  ggplot(aes(x = hour, y = hourly_activity, color = sex)) +
  facet_grid(.~education) +
  geom_point(alpha = 0.1) +
  geom_smooth()+ 
  scale_y_continuous(limits = c(0, 1500))+ 
  labs(title = 'Hourly Acitivity by Education and Sex')
  
```


## Problem 3

Loading Citibike data, we import the four data frames from each of the four months. We add additional variables specifying the year and month of data collection, then combine them all. We also rename the `customer_type` variable for clarity, and we set the month names and weekdays as ordered factors so that they appear in the order as expected during tabling and graphing.  

```{r}

citi = list(
    mutate(
      read_csv('data/Jan 2020 Citi.csv',
               show_col_types = FALSE),
      month = 'January', year = 2020),
    mutate(
      read_csv('data/July 2020 Citi.csv',
               show_col_types = FALSE),
      month = 'July', year = 2020),
    mutate(
      read_csv('data/Jan 2024 Citi.csv',
               show_col_types = FALSE),
      month = 'January', year = 2024),
    mutate(
      read_csv('data/July 2024 Citi.csv',
               show_col_types = FALSE),
      month = 'July', year = 2024)
) |> bind_rows() |> 
  mutate(
    customer_type = member_casual,
    month = factor(month, levels = c('January', 'July')),
    weekdays = factor(weekdays, levels = 
                        c('Monday', 'Tuesday', 'Wednesday', 
                          'Thursday', 'Friday', 'Saturday',
                          'Sunday')),
    year = factor(year)
  )
```

Here we produce a reader-friendly table showing the total number of rides in each combination of year, month, and membership. From the table, we can see that there is far more member ridership than non-member ridership, especially in the month of January 2020. We can also see that there is more ridership in July than in January. This difference is explainable by rider preference for fair weather. 
```{r}
citi |> 
  group_by(year, month, customer_type) |> 
  summarise(ride_count = n()) |> 
  pivot_wider(
    names_from = customer_type,
    values_from = ride_count
  ) |> 
  knitr::kable(
    col.names = c('Year', 'Month', 'Non-member Ride Count', 'Member Ride Count'))
```


Next we investigate the most popular starting stations in July of 2024. 

```{r}
citi |> 
  filter(year == 2024, month == 'July') |> 
  group_by(start_station_name) |> 
  summarise(ride_count = n()) |> 
  arrange(desc(ride_count)) |> 
  head(5) |> 
  knitr::kable(col.names = c('Starting Station', 'Ride Count'))
```
Next we investigate the effect of day of the week, month, and year on median ride duration. Firstly, we can see that median ride duration is shorter in 2024 than it is in 2020. This could be due to rider choice changing, or the availability of additional docking stations, allowing riders to take shorter rides. We can also see that ride duration is typically longer during the month of July, likely due to rider preference for fair weather. Across all months and years, we see a slight uptick in the duration of rides as the week goes on, with the largest median rides occurring on Saturday and Sunday. 
```{r}
citi |> 
  group_by(weekdays, month, year) |> 
  summarise(Median_Duration = median(duration)) |> 
  ggplot(aes(x = month, fill = weekdays, y = Median_Duration))+
  facet_grid(year ~ . ) +
  geom_bar(position = 'dodge', stat = 'identity')+ 
  labs(title = 'Median Citibike Ride Duration')
```

Next we investigate the effect of month, customer type, and bike type ride duration in 2024. Since very few rides are longer than 75 minutes, we filter for rides under 75 minutes long. We then create a four panel graph, faceting by customer type and month. We have used color to show the difference between each bike type. We have normalized each histogram. From the graph, we can see that electric bikes are used less than their human-powered counterparts across month and customer type. This could be due to the the persistent unavailability of electric bikes or consumer choice against electric bikes due to safety, cost, and docking concerns. 
```{r}
citi |> 
  filter(year == 2024, duration < 75) |> 
  ggplot(aes(x = duration, fill = rideable_type))+
  facet_grid(customer_type ~ month) +
  geom_histogram(aes(y = ..density..)) + 
  labs(title = '2024 Electric and Classic Bike Usage')
```


